
---
title: "An empirical example: _Rhynoclemmys_ data."
author: "Daniel R. Miranda-Esquivel"
date: "2020 - 10 - 13"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---


# An empirical example: _Rhynoclemmys_ data.

We read the data, distribution and trees.


```{r}

options(warn=-1)
suppressMessages(library(ggtree))
#~ library(ape)
suppressMessages(library(blepd))
library(ggplot2)
#~ library(ggtree)



## read distributional data

setwd("../data/")

csvFile <- dir(pattern=".csv")


distribution <-          as.matrix(read.table(csvFile,
                                  stringsAsFactors=FALSE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                          )

##qplot(distribution)



## read trees

treeFiles <- dir(pattern=".tre")

treeFiles

RhinoclemmysData <- list()

RhinoclemmysData$distribution <- distribution

RhinoclemmysData$trees  <-    lapply(treeFiles,FUN=read.tree)


distXY <- matrix2XY(distribution)



## plotting

## the trees


par(mfrow=c(2,1))

for( numero in 1:length(treeFiles) ){

plotTree <-  ggtree(RhinoclemmysData$trees[[numero]], ladderize=TRUE,
                    color="black", size=2, linetype="solid") +
             geom_tiplab(size=3, color="black") +
             theme_tree2() +
             ggtitle(treeFiles[numero])


##
print(plotTree)

}



#~ Alternatively, we can plot the trees using ape

#~ for( treeNumber in 1:length(treeFiles) ){
#~ 
#~        plot.phylo(RhinoclemmysData$trees[[treeNumber]],
#~                   main=treeFiles[numero])
#~ 
#~    }





## the distribution


par(mfrow=c(1,1))

plotDistrib <-ggplot(distXY,
               aes(x= Area, y=Terminal), size =11) +
              geom_point() +
               labs(title = "Distributions",
                    y = "",
                    x = "Area")

##
print(plotDistrib)

##cowplot::plot_grid(plotTree, plotDistrib, ncol=2)

# Using faceplot


#~ p2 <- facet_plot(plotTree, panel="dot", 
#~                  data=distXY, 
#~                  geom=geom_point(), 
#~                  mapping=aes(x= distXY$Area,y=distXY$Terminal), 
#~                  size =11, color='red3')


#~ print(p2)

##dev.off()

```


These two trees correspond to different data sources Molecular vs TotalEvidence data, and we want to test whether the data used will have any effect in the areas chosen, or if the area is selected given the terminal(s) branch length.



The branch length histogram shows there are two longer branches,  _R._ _aerolata_ and _R._ _rubida_ and these two are in area A, while area B is the richest.



```{r}

par(mfrow=c(2,1))

qplot(RhinoclemmysData$trees[[1]]$edge.length,bins=10,title=treeFiles[1])


qplot(RhinoclemmysData$trees[[2]]$edge.length,bins=10,title=treeFiles[2])

###dev.off()

```




Fist we calculate the PD value for each tree, we save the values as a table and later we convert the table to a matrix:

```{r}


RhinoclemmysData$tablePD <- lapply(RhinoclemmysData$trees,
                                   FUN=PDindex,
                                   distribution=RhinoclemmysData$distribution,
                                   root=TRUE)



RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))


RhinoclemmysData$tablePDPercentage <- lapply(RhinoclemmysData$trees,
                                             FUN=PDindex,
                                             distribution=RhinoclemmysData$distribution,
                                             percentual=TRUE,
                                             root=TRUE)

RhinoclemmysData$matrixPDPercentage <- as.data.frame(
                                          matrix(
                                              unlist(RhinoclemmysData$tablePDPercentage),
                                              nrow= length(treeFiles),
                                              byrow=TRUE))



```

Now, we test the efect of branch length swapping, terminal and internal branch lengths.


```{r}

for( numero in 1:length(treeFiles) ){
	
	for( modelo in c("simpleswap","allswap","uniform") ){
		
		for( rama in c("terminals","internals") ){
		
		
		
		val <- swapBL(RhinoclemmysData$trees[[numero]],distribution,
		       model=modelo, 
		       branch = rama
		       )
		 
		 cat("\nTree=",treeFiles[numero],"\tModel=",modelo,"\tBranchs swapped=",rama,"\n")      
		print.blepd(val)
		 cat("\n\n\n")
		 
	 }
	
  }

}


```

For both trees, the terminal branch length is critical in the decision taken, if we swap all terminal branch lengths (model=_allswap_), or if we replace them with a uniform distribution (model=_uniform_), the area selected changes from A to B.

As the terminal branch lengths are distributed unequally, we might suspect that the results depend heavily on the longest branches that inhabit the area A.


```{r}


## Testing the effect if we reduce the branch lenght form the actual value to 0.0 using evalTerminal {lower}

testEvalTerminalALL <- evalTerminal(RhinoclemmysData$trees[[1]],distribution,tipToEval = "all",approach = "lower")

print.multiBlepd(testEvalTerminalALL)


## Testing the effect if we increase the branch lenght form the actual value to the sum of all branch length  using evalTerminal {upper}


testEvalTerminalALL <- evalTerminal(RhinoclemmysData$trees[[1]],distribution,tipToEval = "all",approach = "upper")

print.multiBlepd(testEvalTerminalALL)


```




The area selected depends heavily in the branch length of _R._ _aerolata_ and _R._ _rubida_, if the branch length is just a little shorter, the area selected will change from area A to area B.


To test whether internal branches have more impact than terminal branches, we can use the function evalTerminalvsInternal.


```{r}


#~  ## calculate values

data1 <- evalTerminalvsInternal(RhinoclemmysData$trees[[1]],distribution,nTimes=100,branch = "terminals")
data2 <- evalTerminalvsInternal(RhinoclemmysData$trees[[1]],distribution,nTimes=100,branch = "internals")



for( i in 1:3 ){
	
	cat("name ",i,"\n")
	print.blepd(data1[[i]])
	print.blepd(data2[[i]])

}




```
