---
title: "An empirical example: _Rhynoclemmys_ data."
author: "Daniel R. Miranda-Esquivel"
date: "2020 - 11 - 16"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
bibliography: "/home/rafael/disco2/uis/proyectosDRME/indices/mainPaper/bib/blepd.bib"
---




We read the distribution and the tree.



```{r eval=TRUE, echo=TRUE}

## for reproductibility purposes

set.seed(121)


options(warn=-1)
suppressMessages(library(ggtree))
#~ library(ape)
suppressMessages(library(blepd))
library(ggplot2)
#~ library(ggtree)


## Version

cat("Analyses made with blepd version:",unlist(packageVersion("blepd")))



## Create an object to place the distribution and the tree

RhinoclemmysData <- list()


## Read data

## distribution is a labeled csv file, areas by terminals 


setwd("../data/")

csvFile <- dir(pattern=".csv")


### the functions use a matrix object for the distributions

RhinoclemmysData$distribution <-  as.matrix(read.table(csvFile,
                                  stringsAsFactors=TRUE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                                  )


print(t(RhinoclemmysData$distribution))



## tree(s) in nexus or newick format

treeFiles <- dir(pattern=".tre")

##treeFiles


RhinoclemmysData$tree  <-    read.tree(treeFiles)


## name of tree(s)

treeFiles <- gsub(".tre","",treeFiles)





## Plotting


#par(mfrow=c(2,1))


## The tree

## using ggtree

RhinoclemmysData$tree <- reorder(RhinoclemmysData$tree, order = "cladewise")

plotTree <-  ggtree(RhinoclemmysData$tree, ladderize=TRUE,
                    color="black", size=0.51, linetype="solid") +
             geom_tiplab(size=4, color="black") +
             xlim(0,0.030) +
             theme_tree2() +
             ggtitle(treeFiles[1])



##print(plotTree)


#~ Alternatively, we can plot the trees using APE
##
## plot.phylo(RhinoclemmysData$tree)
##
####nodelabels()
####tiplabels()




## to plot the distribution, we must transform it into a data.frame object
## and reorder the data.frame following the names on the trees

distXY <- matrix2XY(RhinoclemmysData$distribution)

terminals <- colnames(RhinoclemmysData$distribution)

realOrder <- match(terminals,RhinoclemmysData$tree$tip.label)

equivalencias <- data.frame(terminals,realOrder)


dXY2 <-     distXY

for(cambiar in terminals){

    dXY2$Terminal[distXY$Terminal == cambiar]     <-
    equivalencias$realOrder[equivalencias$terminals==cambiar]

}



distGraficar <- distXY
#~ distGraficar <- dXY2


## plot using ggplot

plotDistrib <- ggplot(distGraficar,
                      aes(x= Area, y=Terminal), size =30) +
               geom_point(shape=19, fill="white", color="darkgrey", size=4) +
               labs(title = "Distribution",
                    y = "",
                    x = "Area") +
               theme(axis.line=element_blank(),
					# axis.text.y=element_blank(), 
					 axis.ticks=element_blank(), 
					# axis.title.y=element_blank(), 
					 legend.position="none", 
					 panel.background=element_blank(), 
					 panel.border=element_blank(), 
					 panel.grid.major=element_blank(), 
					 panel.grid.minor=element_blank(), 
					 plot.background=element_blank()
					 ) 


##distGraficar <- dXY2


#~ 

cowplot::plot_grid(plotTree,plotDistrib, ncol=1)

#~ plot(plotTree)

#~ plot(plotDistrib)

#~ print(t(RhinoclemmysData$distribution))


##dev.off()

```

We want to test whether the branch lengths will have any effect in the
area(s) chosen for conservation using a Phylogenetic Diversity (PD)
analysis.

The topology corresponds to a Total Evidence analysis for the
_Rhynoclemmys_ genus, from @alarcon2020a, and the distribution is
taken and modified from @le2008.




```{r eval=TRUE, echo=TRUE}

#par(mfrow=c(2,1))


#par(mfrow=c(2,1))

a  <- qplot(RhinoclemmysData$tree$edge.length,
            bins=12,
            main=treeFiles,
            xlab="Branch length: internals and terminals") +
            xlim(c(0.001,0.025)) +
            ylim(c(0,5))


terminals <- RhinoclemmysData$tree$edge[,2] < 1+length(RhinoclemmysData$tree$tip.label)


b  <- qplot(RhinoclemmysData$tree$edge.length[terminals],
            bins=12,
            #main=treeFiles,
            xlab="Branch length: terminals")  +
            xlim(c(0.001,0.025)) +
            ylim(c(0,5))


c  <- qplot(RhinoclemmysData$tree$edge.length[!terminals],
            bins=12,
            #main=treeFiles,
            xlab="Branch length: internals")  +
            xlim(c(0.001,0.025)) +
            ylim(c(0,5))


#~ 

cowplot::plot_grid(a, b, c, nrow=3)

###dev.off()

```

The branch length histograms and the tree plot, show the internal length 
branches are similar, and different to terminals'; there are two longer 
branches, _R._ _aerolata_ (inhabiting Al) and _R._ _rubida_ (Pl), while 
Al and Cho are the richest areas.


Fist we calculate the PD value for the tree, we save the values as a
table and later we convert the table to a matrix.


```{r eval=TRUE, echo=TRUE}


RhinoclemmysData$tablePD <- PDindex(RhinoclemmysData$tree,
                                    distribution=RhinoclemmysData$distribution,
                                    root=TRUE)



RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))



colnames(RhinoclemmysData$matrixPD) <- row.names(RhinoclemmysData$distribution)

row.names(RhinoclemmysData$matrixPD) <- "PD value"

print(t(sort(RhinoclemmysData$matrixPD,decreasing = TRUE)))

```


The  highest PD is for area Cho, followed by Al, the two richest areas.


Now, we test the efect of branch length on these values, swapping terminal and internal branch lengths.


```{r eval=TRUE, echo=TRUE}


 for( modelo in c("simpleswap","allswap","uniform") ){

        for( rama in c("terminals","internals") ){

            val <- swapBL(RhinoclemmysData$tree,
                          RhinoclemmysData$distribution,
                          model = modelo,
                         branch = rama
               )

         cat("\n\tTree=",treeFiles,"\n\tModel=",modelo,"\n\tBranchs swapped=",rama,"\n")

        print.blepd(val)

         cat("\n\n\n")

     }

  }



```

The terminal branch length is critical in the decision taken, if we
swap all terminal branch lengths (model=_allswap_), or if we replace
them with a uniform distribution (model=_uniform_), the area selected
might change from Cho to Al. As the terminal branch lengths are distributed
unequally, we might suspect that the results could depend  on the
longest branches that inhabit the area Al.

But first, let us see if the number of replicates has any effect.

```{r eval=TRUE, echo=TRUE}

for(repetir in 1:4){

            val <- swapBL( RhinoclemmysData$tree,
                           RhinoclemmysData$distribution,
                           model = "allswap",
                           branch = "terminals",
                           nTimes = 10**repetir
                         )
                         
            print.blepd(val)

}
```

Roughly speaking from 1000 on the results are alike, and the
largest difference is in 10 -- 100. As a rule of thumb, we must use at
least 100 replicates, but 1000 will be better.



now, let us see if the possible difference in results could be assigned to both of our longest branches.

```{r eval=TRUE, echo=TRUE}
## Testing the effect if we reduce the branch lenght form the actual
## value to 0.0 using evalTerminal {lower}

testEvalTerminalALL <- evalTerminal(RhinoclemmysData$tree,
                                    RhinoclemmysData$distribution,
                                    tipToEval = "all",
                                    approach = "lower")


print.multiBlepd(testEvalTerminalALL)


## Testing the effect if we increase the branch lenght form the actual
## value to the sum of all branch length using evalTerminal {upper}


testEvalTerminalALL <- evalTerminal(RhinoclemmysData$tree,
                                    RhinoclemmysData$distribution,
                                    tipToEval = "all",
                                    approach = "upper")

print.multiBlepd(testEvalTerminalALL)

```


The area selected depends heavily in the branch length of _R._ _nasuta_. If the terminal branch length is just a little shorter (-0.0047%), the area selected will change from Cho to Al. But, if the terminal branch length for _R._ _diademata_ and _R._ _punctuluaria_ are 0.18 or 0.12 % larger, the area selected will change from Cho to Nsa or An-As.


# Literature cited
