---
title: "An empirical example: _Rhinoclemmys_ data"
author: "Daniel R. Miranda-Esquivel"
date: "2025 / 01 / 05"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
bibliography: "../../bib/blepd002.bib"

---


I conducted a phylogenetic diversity (PD) analysis for the _Rhinoclemmys_ genus to evaluate the effect of branch length on the chosen conservation area(s). The topology used corresponds to a total evidence analysis from [@alarcon2020a], while the distribution was modified from [@le2008]. 



# Read the data: the distribution, and the tree


```{r eval=TRUE, echo=TRUE}

## For reproductibility purposes

set.seed(121)


options(warn=-1)
library(ggplot2)
suppressMessages(library(ggtree))

suppressMessages(library(blepd))

suppressMessages(library(phytools))

## library(ggtree) 

##
## If you want to install ggtree. 
##
### https://bioconductor.org/packages/release/bioc/html/ggtree.html
##
####
#  if (!require("BiocManager", quietly = TRUE))
#      install.packages("BiocManager")
#  BiocManager::install("ggtree")

## Version

cat("Analyses made with blepd version:",unlist(packageVersion("blepd")))

##
## The tree file (in Newick format) and the distribution file (in CSV format) 
## can be found in the data directory or as a data set within the package. 
##

## Assuming data are read from files, create an object to store the 
## distribution and the tree.

RhinoclemmysData <- list()

## Read data

## The distribution is a labeled CSV file with areas as rows and terminals as 
## columns.

#setwd("./csv/")

## getwd()

csvFile <- list.files(pattern="csv")

## csvFile

### The functions use a matrix object for the distributions

RhinoclemmysData$distribution <-  as.matrix(read.table(csvFile,
                                  stringsAsFactors=FALSE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                                  )


print(t(RhinoclemmysData$distribution))

##
## Tree(s) in nexus or newick format
##
## setwd("../tree/")
##

treeFiles <- list.files(pattern=".tre")

##treeFiles

RhinoclemmysData$tree  <-    read.tree(treeFiles)


## name of tree(s) without .tre

treeFiles <- gsub(".tre","",treeFiles)


##getwd()

## You might need to save the object as an .Rda file.

##
save(RhinoclemmysData, file = "../data/RhinoclemmysData.rda")



## Plotting:

## The tree

## using ggtree

# RhinoclemmysData$tree <- reorder(RhinoclemmysData$tree, order = "cladewise")

plotTree <-  ggtree(RhinoclemmysData$tree, ladderize=TRUE,
                    color="black", size=0.51, linetype="solid") +
             geom_tiplab(size=4, color="black") +
             xlim(0,0.030) +
             theme_tree2() +
             ggtitle(treeFiles[1])


##pdf("RhinoclemmysTopology.pdf", onefile = TRUE)

print(plotTree)

##dev.off()


#~ Alternatively, we can plot the tree using APE


plot.phylo(RhinoclemmysData$tree)


# The number of the nodes / tips 
#
nodelabels()
####
tiplabels()
 
plot(plotTree)
 
print(t(RhinoclemmysData$distribution))



```

#Plot the branch lengths.

```{r eval=TRUE, echo=TRUE}

library(ggplot2)

##par(mfrow=c(2,1))

datos <- RhinoclemmysData$tree$edge.length
data <- as.data.frame(datos)

a <- ggplot2::ggplot(data, aes(x=datos)) +
     geom_histogram(bins = 12) +
     labs(title = treeFiles, x = "Branch length: internals and terminals", y = "Frequency") +
     xlim(c(0.001, 0.025)) +
     ylim(c(0, 5))
  
terminals <- RhinoclemmysData$tree$edge[,2] < 1 + length(RhinoclemmysData$tree$tip.label)

datos <- RhinoclemmysData$tree$edge.length[terminals]
data <- as.data.frame(datos)

b <- ggplot(data, aes(x=datos)) +
  geom_histogram(bins = 12) +
  labs(x = "Branch length: terminals", y = "Frequency") +
  xlim(c(0.001, 0.025)) +
  ylim(c(0, 5))


  
datos <- RhinoclemmysData$tree$edge.length[!terminals]
data <- as.data.frame(datos)

c <- ggplot(data, aes(x=datos)) +
  geom_histogram(bins = 12) +
  labs(x = "Branch length: internals", y = "Frequency") +
  xlim(c(0.001, 0.025)) +
  ylim(c(0, 5))


##pdf("RhinoclemmysBranchLength.pdf", onefile = TRUE)

cowplot::plot_grid(a, b, c, nrow=3)

##dev.off()


#
# cat("Terminals, with BL larger than 0.02 = ",
# findTerminalgivenLength(RhinoclemmysData$tree,0.02))
#



```

The branch length histograms and the tree plot showed that the internal length 
branches were similar among each other, and different to terminals'; there were 
two longer branches (larger than 0.02), _R._ _nasuta_ (inhabiting Cho) and 
_R._ _rubida_ (Pl), whereas Al and Cho were the richest.


# PD values

First, I calculated the PD value for the areas, given this tree.


```{r eval=TRUE, echo=TRUE}

RhinoclemmysData$tablePD <- PDindex(RhinoclemmysData$tree,
                              distribution=RhinoclemmysData$distribution,
                                    root=TRUE,
                                    percentual = TRUE)

RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))


colnames(RhinoclemmysData$matrixPD) <- row.names(RhinoclemmysData$distribution)

row.names(RhinoclemmysData$matrixPD) <- "PD value"

 new <- (t(RhinoclemmysData$matrixPD))

 
values <-  new[order(new[,1])]
names  <-  rownames(new)[order(new[,1])]

print(as.data.frame(paste(names,values)))

 
```

The highest PD was for area Cho, followed by Al, the two richest areas. However, the difference in PD values was not determined by species richness alone, but also by the branch lengths of the species within each region.


#Effect of branch lengths in the PD

I tested the effect of branch length on PD values by swapping terminal and/or internal branch lengths using three different models.

```{r eval=TRUE, echo=TRUE}


 for( modelo in c("simpleswap","allswap","uniform") ){

        for( rama in c("terminals","internals","all") ){

            val <- swapBL(RhinoclemmysData$tree,
                          RhinoclemmysData$distribution,
                          model = modelo,
                          branch = rama
               )

         cat( "\n\t Tree=",treeFiles,"\n\t Model=",modelo,
             "\n\t Branchs swapped=",rama,"\n" )

        printSwapBL(val)

         cat( "\n\n" )

     }

  }

```


I determined that terminal branch lengths were crucial in influencing the chosen area. When I swapped all terminal branch lengths (model = "allswap") or replaced them with a uniform distribution (model = "uniform"), the optimal area shifted from Cho to Al. Recognizing the uneven distribution of terminal branch lengths, I suspected that the longest branches within the Cho and/or Al areas were driving these results.


I first investigated the impact of the number of replicates on the results.

```{r eval=TRUE, echo=TRUE}

for(repetir in 1:4){

            val <- swapBL( RhinoclemmysData$tree,
                           RhinoclemmysData$distribution,
                           model = "allswap",
                           branch = "all",
                           nTimes = 10**repetir
                         )
                         
            printSwapBL(val)

}
```

I found that results stabilized after approximately 1000 replicates, with most differences arising from using 10 or 100 replicates. As a general guideline, at least 100 replicates are necessary, but 1000 are preferable. 

I investigated whether the observed differences in results could be attributed to the lengths of the longest branches.



```{r testEvalBranch, eval=TRUE, echo=TRUE}

## we can use a single command (*evalBranch*)

AllTerminals <- evalBranch(tree=RhinoclemmysData$tree,
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="terminals", 
                           approach="all",
                           printNames=TRUE)

printEvalBranch(AllTerminals)

AllInternals <- evalBranch(tree=RhinoclemmysData$tree,
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="internals", 
                           approach="all",
                           printNames=TRUE)

printEvalBranch(AllInternals)

```

I found that branch length significantly influenced the selected area. While modifying internal branches had minimal to no effect, altering terminal branch lengths might have an impact on area selection. Specifically, a 46% decrease in the branch length of _R. nasuta_ or an 86% increase in the branch length of _R. areolata shifted the optimal area from Cho to Al. Similarly, a 59% increase in the branch length of _R. rubida or a 116% increase in the branch length of _R. pulcherrima resulted in a shift from Cho to Pl.


As an additional exercise, I investigated the impact of ultrametricity. I hypothesized that if the tree were ultrametric, terminal sister pairs would exhibit similar behavior, with differences primarily attributable to the underlying distribution.



```{r testEvalBranchUltra, eval=TRUE, echo=TRUE}

## now let's see the impact of ultra-metricity

AllTerminalsU <- evalBranch(tree=force.ultrametric(RhinoclemmysData$tree),
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="terminals", 
                           approach="all",
                           printNames=TRUE)

cat("Ultrametric tree\n")

printEvalBranch(AllTerminalsU)


printEvalBranch(AllTerminals)
                    
AllInternalsU <- evalBranch(tree=force.ultrametric(RhinoclemmysData$tree),
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="internals", 
                           approach="all",
                           redondeo=3,
                           printNames=TRUE)

printEvalBranch(AllInternalsU)

printEvalBranch(AllInternals)



```

These findings suggest that decisions to conserve solely area Cho are not misguided. Areas Al, and potentially Pl, should not be considered as essential components of the conservation range.



# Literature cited
