---
title: "An empirical example: _Rhynoclemmys_ data"
author: "Daniel R. Miranda-Esquivel"
date: "2022/03/11"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
bibliography: "/home/rafael/disco2/proyectosDRME/indices/paperBlepd/bib/blepd.bib"
---


We want to test whether the branch lengths have any effect in the 
area(s) chosen for conservation, based on a Phylogenetic Diversity 
(PD) analysis for the _Rhynoclemmys_ genus. The topology corresponds to 
a Total Evidence analysis from @alarcon2020a, and the distribution is modified 
from @le2008.



# We read the distribution and the tree


```{r eval=TRUE, echo=TRUE}

## For reproductibility purposes

set.seed(121)


options(warn=-1)
suppressMessages(library(ggtree))
#~ library(ape)
suppressMessages(library(blepd))
library(ggplot2)
#~ library(ggtree)


## Version

cat("Analyses made with blepd version:",unlist(packageVersion("blepd")))



## Create an object to place the distribution and the tree

RhinoclemmysData <- list()


## Read data

## distribution is a labeled csv file, areas by terminals 


#setwd("./csv/")

##

getwd()


csvFile <- list.files(pattern="csv")

##csvFile


### the functions use a matrix object for the distributions

RhinoclemmysData$distribution <-  as.matrix(read.table(csvFile,
                                  stringsAsFactors=FALSE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                                  )


print(t(RhinoclemmysData$distribution))



## tree(s) in nexus or newick format

##setwd("../tree/")

treeFiles <- list.files(pattern=".tre")

##treeFiles


RhinoclemmysData$tree  <-    read.tree(treeFiles)


## name of tree(s)

treeFiles <- gsub(".tre","",treeFiles)





## Plotting


#par(mfrow=c(2,1))


## The tree

## using ggtree

# RhinoclemmysData$tree <- reorder(RhinoclemmysData$tree, order = "cladewise")

plotTree <-  ggtree(RhinoclemmysData$tree, ladderize=TRUE,
                    color="black", size=0.51, linetype="solid") +
             geom_tiplab(size=4, color="black") +
             xlim(0,0.030) +
             theme_tree2() +
             ggtitle(treeFiles[1])



##print(plotTree)


#~ Alternatively, we can plot the trees using APE
##
## plot.phylo(RhinoclemmysData$tree)
##
####nodelabels()
####tiplabels()



## to plot the distribution, we must transform it into a data.frame 
## object 


distXY <- matrix2XY(RhinoclemmysData$distribution)


## We could reorder the data.frame following the names on the trees


terminals <- colnames(RhinoclemmysData$distribution)

realOrder <- match(terminals,RhinoclemmysData$tree$tip.label)

equivalencias <- data.frame(terminals,realOrder)

dXY2 <-  distXY

for(cambiar in terminals){

    dXY2$Terminal[distXY$Terminal == cambiar]     <-
    equivalencias$realOrder[equivalencias$terminals==cambiar]

}



distGraficar <- distXY

#~ distGraficar <- dXY2


## plot using ggplot

plotDistrib <- ggplot(distGraficar,
                      aes(x= Area, y=Terminal), size =30) +
               geom_point(shape=19, fill="white", 
                          color="darkgrey", 
                          size=4) +
               labs(title = "Distribution",
                    y = "",
                    x = "Area") +
               theme(axis.line=element_blank(),
		    			# axis.text.y=element_blank(), 
					 axis.ticks=element_blank(), 
			    		# axis.title.y=element_blank(), 
					 legend.position="none", 
					 panel.background=element_blank(), 
					 panel.border=element_blank(), 
					 panel.grid.major=element_blank(), 
					 panel.grid.minor=element_blank(), 
					 plot.background=element_blank()
					 ) 




#~ 

cowplot::plot_grid(plotTree,plotDistrib, ncol=1)

#~ plot(plotTree)

#~ plot(plotDistrib)

#~ print(t(RhinoclemmysData$distribution))


##dev.off()

```

# Now we plot the branch lengths

```{r eval=TRUE, echo=TRUE}


#par(mfrow=c(2,1))

a  <- qplot(RhinoclemmysData$tree$edge.length,
            bins=12,
            main=treeFiles,
            xlab="Branch length: internals and terminals") +
            xlim(c(0.001,0.025)) +
            ylim(c(0,5))


terminals <- RhinoclemmysData$tree$edge[,2] < 1 + length(RhinoclemmysData$tree$tip.label)


b  <- qplot(RhinoclemmysData$tree$edge.length[terminals],
            bins=12,
            #main=treeFiles,
            xlab="Branch length: terminals")  +
            xlim(c(0.001,0.025)) +
            ylim(c(0,5))


c  <- qplot(RhinoclemmysData$tree$edge.length[!terminals],
            bins=12,
            #main=treeFiles,
            xlab="Branch length: internals")  +
            xlim(c(0.001,0.025)) +
            ylim(c(0,5))

## 

cowplot::plot_grid(a, b, c, nrow=3)

##
#
#cat("Terminals, with BL larger than 0.02 = ",findTerminalgivenLength(RhinoclemmysData$tree,0.02))
#
##


###dev.off()

```

The branch length histograms and the tree plot show that the internal length 
branches are similar among each other, and different to terminals'; there are 
two longer branches (larger than 0.02), _R._ _nasuta_ (inhabiting Cho) and 
_R._ _rubida_ (Pl), while the areas Al and Cho are the richest.


# PD values

First, we calculate the PD value for the areas, given this tree.


```{r eval=TRUE, echo=TRUE}


RhinoclemmysData$tablePD <- PDindex(RhinoclemmysData$tree,
                                    distribution=RhinoclemmysData$distribution,
                                    root=TRUE,
                                    percentual = TRUE)



RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))



colnames(RhinoclemmysData$matrixPD) <- row.names(RhinoclemmysData$distribution)

row.names(RhinoclemmysData$matrixPD) <- "PD value"

print(t(sort(RhinoclemmysData$matrixPD,decreasing = TRUE)))

```


The  highest PD is for area Cho, followed by Al, the two richest areas, 
and the difference in PD value is not given by the richness but the 
species inhabiting each area.

#Effect of branch lengths in the PD

We test the effect of the branch length on the PD values by swapping 
terminal and/or internal branch lengths, using the three available models.


```{r eval=TRUE, echo=TRUE}


 for( modelo in c("simpleswap","allswap","uniform") ){

        for( rama in c("terminals","internals","all") ){

            val <- swapBL(RhinoclemmysData$tree,
                          RhinoclemmysData$distribution,
                          model = modelo,
                         branch = rama
               )

         cat( "\n\t Tree=",treeFiles,"\n\t Model=",modelo,
             "\n\t Branchs swapped=",rama,"\n" )

        print.blepd(val)

         cat( "\n\n\n" )

     }

  }



```

The terminal branch length is critical in the decision taken, if we 
swap all terminal branch lengths (model= "allswap"), or if we replace 
them with a uniform distribution (model= "uniform"), the area selected 
might change from Cho to Al. 

As the terminal branch lengths are distributed unequally, we might 
suspect that the results could depend on the longest branches that 
inhabit the areas Cho/Al.

But first, we test if the number of replicates has any effect.

```{r eval=TRUE, echo=TRUE}

for(repetir in 1:5){

            val <- swapBL( RhinoclemmysData$tree,
                           RhinoclemmysData$distribution,
                           model = "allswap",
                           branch = "terminals",
                           nTimes = 10**repetir
                         )
                         
            print.blepd(val)

}
```

Roughly speaking, from 1000 on the results are alike, and the largest 
difference is using 10 or 100 replicates. As a rule of thumb, we must 
use at least 100 replicates, but 1000 will be better.

Now, let us see if the possible difference in results could be assigned 
to the longest branches.

```{r eval=TRUE, echo=TRUE}
## Testing the effect if we change the branch length from the actual
## value to 0.0. We use evalTerminal(approach = "lower")

testEvalTerminalALL <- evalTerminal( RhinoclemmysData$tree,
                                     RhinoclemmysData$distribution,
                                     tipToEval = "all",
                                     approach = "lower" )


print.multiBlepd( testEvalTerminalALL )


## Testing the effect if we change the branch length from the actual
## value to the sum of all branch lengths. We use evalTerminal(approach = "upper")


testEvalTerminalALL <- evalTerminal( RhinoclemmysData$tree,
                                     RhinoclemmysData$distribution,
                                     tipToEval = "all",
                                     approach = "upper" )

print.multiBlepd( testEvalTerminalALL )

```


The area selected depends on the branch length. If the terminal branch 
length of _R._ _nasuta_ is -47.18\% shorter, or if the terminal branch 
length is 84.95% larger for _R._ _areolata_, the area selected will 
change from Cho to Al, and if the terminal branch length of _R._ _rubida_ 
is 61.32\% larger, the area selected will change from Cho to Pl.


# Literature cited
