---
title: "An empirical example: _Rhinoclemmys_ data"
author: "Daniel R. Miranda-Esquivel"
date: "2024 / 07 / 20"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
bibliography: "/home/rafael/bib/blepd.bib"
---

To evaluate the effect of branch length on the area(s) chosen for conservation, I conduct a Phylogenetic Diversity (PD) analysis for the _Rhinoclemmys_ genus. The topology I will use corresponds to a Total Evidence analysis from @alarcon2020a, and the distribution will be modified from @le2008.



# Read the data: the distribution, and the tree


```{r eval=TRUE, echo=TRUE}

## For reproductibility purposes

set.seed(121)


options(warn=-1)
library(ggplot2)
suppressMessages(library(ggtree))

suppressMessages(library(blepd))

suppressMessages(library(phytools))

#~ library(ggtree) 
##
## Just in case, if you want to install ggtree. 
##
### https://bioconductor.org/packages/release/bioc/html/ggtree.html
##
####
#  if (!require("BiocManager", quietly = TRUE))
#      install.packages("BiocManager")
#  BiocManager::install("ggtree")

## Version

cat("Analyses made with blepd version:",unlist(packageVersion("blepd")))


## Create an object to place the distribution and the tree

RhinoclemmysData <- list()

## Read data

## distribution is a labeled csv file, areas by terminals 

#setwd("./csv/")

## getwd()

csvFile <- list.files(pattern="csv")

## csvFile

### the functions use a matrix object for the distributions

RhinoclemmysData$distribution <-  as.matrix(read.table(csvFile,
                                  stringsAsFactors=FALSE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                                  )


print(t(RhinoclemmysData$distribution))


## tree(s) in nexus or newick format

##setwd("../tree/")

treeFiles <- list.files(pattern=".tre")

##treeFiles

RhinoclemmysData$tree  <-    read.tree(treeFiles)


## name of tree(s)

treeFiles <- gsub(".tre","",treeFiles)



## Plotting:

## The tree

## using ggtree

# RhinoclemmysData$tree <- reorder(RhinoclemmysData$tree, order = "cladewise")

plotTree <-  ggtree(RhinoclemmysData$tree, ladderize=TRUE,
                    color="black", size=0.51, linetype="solid") +
             geom_tiplab(size=4, color="black") +
             xlim(0,0.030) +
             theme_tree2() +
             ggtitle(treeFiles[1])




print(plotTree)


#~ Alternatively, we can plot the trees using APE


plot.phylo(RhinoclemmysData$tree)


####nodelabels()
####tiplabels()
 
plot(plotTree)
 
print(t(RhinoclemmysData$distribution))



```

# Now we plot the branch lengths

```{r eval=TRUE, echo=TRUE}

library(ggplot2)

##par(mfrow=c(2,1))

datos <- RhinoclemmysData$tree$edge.length
data <- as.data.frame(datos)

a <- ggplot2::ggplot(data, aes(x=datos)) +
     geom_histogram(bins = 12) +
     labs(title = treeFiles, x = "Branch length: internals and terminals", y = "Frequency") +
     xlim(c(0.001, 0.025)) +
     ylim(c(0, 5))
  
terminals <- RhinoclemmysData$tree$edge[,2] < 1 + length(RhinoclemmysData$tree$tip.label)

datos <- RhinoclemmysData$tree$edge.length[terminals]
data <- as.data.frame(datos)

b <- ggplot(data, aes(x=datos)) +
  geom_histogram(bins = 12) +
  labs(x = "Branch length: terminals", y = "Frequency") +
  xlim(c(0.001, 0.025)) +
  ylim(c(0, 5))


  
datos <- RhinoclemmysData$tree$edge.length[!terminals]
data <- as.data.frame(datos)

c <- ggplot(data, aes(x=datos)) +
  geom_histogram(bins = 12) +
  labs(x = "Branch length: internals", y = "Frequency") +
  xlim(c(0.001, 0.025)) +
  ylim(c(0, 5))




cowplot::plot_grid(a, b, c, nrow=3)

#
#  cat("Terminals, with BL larger than 0.02 = ",
# findTerminalgivenLength(RhinoclemmysData$tree,0.02))
#



```

The branch length histograms and the tree plot show that the internal length 
branches are similar among each other, and different to terminals'; there are 
two longer branches (larger than 0.02), _R._ _nasuta_ (inhabiting Cho) and 
_R._ _rubida_ (Pl), whereas Al and Cho are the richest.


# PD values

First, I calculate the PD value for the areas, given this tree.


```{r eval=TRUE, echo=TRUE}

RhinoclemmysData$tablePD <- PDindex(RhinoclemmysData$tree,
                              distribution=RhinoclemmysData$distribution,
                                    root=TRUE,
                                    percentual = TRUE)

RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))


colnames(RhinoclemmysData$matrixPD) <- row.names(RhinoclemmysData$distribution)

row.names(RhinoclemmysData$matrixPD) <- "PD value"

 new <- (t(RhinoclemmysData$matrixPD))

 
values <-  new[order(new[,1])]
names  <-  rownames(new)[order(new[,1])]

print(as.data.frame(paste(names,values)))

 
```


The  highest PD is for area Cho, followed by Al, the two richest areas, 
and the difference in PD value is not given by the richness but the 
species in each region.

#Effect of branch lengths in the PD

We test the effect of the branch length on the PD values by swapping 
terminal and/or internal branch lengths using the three available models.


```{r eval=TRUE, echo=TRUE}


 for( modelo in c("simpleswap","allswap","uniform") ){

        for( rama in c("terminals","internals","all") ){

            val <- swapBL(RhinoclemmysData$tree,
                          RhinoclemmysData$distribution,
                          model = modelo,
                          branch = rama
               )

         cat( "\n\t Tree=",treeFiles,"\n\t Model=",modelo,
             "\n\t Branchs swapped=",rama,"\n" )

        printSwapBL(val)

         cat( "\n\n" )

     }

  }

```


The terminal branch length is critical in the decision taken, if we 
swap all terminal branch lengths (model= "allswap"), or if we replace 
them with a uniform distribution (model= "uniform"), the area selected 
change from Cho to Al. 

As the terminal branch lengths are distributed unequally, we might 
suspect that the results could depend on the longest branches that 
inhabit Cho/Al areas.

First, we tested whether the number of replicates had any effect.

```{r eval=TRUE, echo=TRUE}

for(repetir in 1:2){

            val <- swapBL( RhinoclemmysData$tree,
                           RhinoclemmysData$distribution,
                           model = "allswap",
                           branch = "terminals",
                           nTimes = 10**repetir
                         )
                         
            printSwapBL(val)

}
```

Roughly speaking, from 1000 on the results are alike, and the largest 
The difference was the use of 10 or 100 replicates. As a rule of thumb, we must 
use at least 100 replicates, but 1000 replicates would be better.

Now, let us see if the difference in results could be assigned to the longest 
branches (or not).

```{r testEvalBranch, eval=TRUE, echo=TRUE}

## we can use a single command (*evalBranch*)

AllTerminals <- evalBranch(tree=RhinoclemmysData$tree,
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="terminals", 
                           approach="all",
                           printNames=TRUE)

printEvalBranch(AllTerminals)

AllInternals <- evalBranch(tree=RhinoclemmysData$tree,
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="internals", 
                           approach="all",
                           printNames=TRUE)

printEvalBranch(AllInternals)

```

The selected area depended on the branch length. While the internal terminals had 
almost no impact, if the terminal branch length of _R._ _nasuta_ is -47.18\% 
shorter, or if the terminal branch length is 84.95% larger for _R._ _areolata_, 
the area selected will change from Cho to Al, and if the terminal branch length 
of _R._ _rubida_ is 61.32\% larger, and the area selected changes from Cho to Pl.


If the tree is ultrametric, the pair of sister pairs will behave alike, and the 
difference will be given by the distribution.


```{r testEvalBranchUltra, eval=TRUE, echo=TRUE}

## now let's see the impact of ultra-metricity

AllTerminalsU <- evalBranch(tree=force.ultrametric(RhinoclemmysData$tree),
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="terminals", 
                           approach="all",
                           printNames=TRUE)

cat("Ultrametric tree\n")

printEvalBranch(AllTerminalsU)


printEvalBranch(AllTerminals)
                    
AllInternalsU <- evalBranch(tree=force.ultrametric(RhinoclemmysData$tree),
                           distribution=RhinoclemmysData$distribution,
                           branchToEval="internals", 
                           approach="all",
                           redondeo=3,
                           printNames=TRUE)

printEvalBranch(AllInternalsU)

printEvalBranch(AllInternals)



```



# Literature cited
