# An empirical example: _Rhynoclemmys_ data.

We read the data sets: distribution and trees.


```{r}

library(ape)
library(blepd)
library(ggplot2)
library(ggtree)


## read distributional data

setwd("../data/")

csvFile <- dir(pattern=".csv")


distribution <-          as.matrix(read.table(csvFile,
                                  stringsAsFactors=FALSE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                          )

##qplot(distribution)



## trees

treeFiles <- dir(pattern=".tre")

treeFiles


RhinoclemmysData <- list()

RhinoclemmysData$distribution <- distribution

RhinoclemmysData$trees  <-    lapply(treeFiles,FUN=read.tree)

RhinoclemmysData$trees[[1]]$tip.label


### reorder distribution to plot it

distXY <- matrix2XY(distribution)

#qplot(distXY)


## order the distribution object

ordenados <- match(RhinoclemmysData$trees[[1]]$tip.label,distXY$Terminal)


distXY <- distXY[ordenados,]

## terminals as factors to plot the same orden in tree and distribution

distXY$Terminal <- factor(distXY$Terminal, levels = distXY$Terminal)


##
###
## https://stackoverflow.com/questions/38131596/ggplot2-geom-bar-how-to-keep-order-of-data-frame
###
##



## plotting

## the tree

plotTree <-  ggtree(RhinoclemmysData$trees[[1]], ladderize=TRUE,
                    color="black", size=1, linetype="solid") +
             geom_tiplab(size=6, color="black") +
             theme_tree2() +
             ggtitle(treeFiles[1])


##
print(plotTree)



## the distribution

plotDistrib <-ggplot(distXY,
               aes(x= Area, y=Terminal), size =11) +
              geom_point() +
               labs(title = "Distributions",
                    y = "",
                    x = "Area")

##
print(plotDistrib)

cowplot::plot_grid(plotTree, plotDistrib, ncol=2)



dev.off()

```

These two trees correspond to different data sources Molecular vs TotalEvidence data, and we want to test whether the data used will have any effect in the areas chosen.


Fist we calculate the PD value for each tree, we save the values as a table and later we convert the table to a matrix:

```{r}


RhinoclemmysData$tablePD <- lapply(RhinoclemmysData$trees,
                                   FUN=PDindex,
                                   distribution=RhinoclemmysData$distribution,
                                   root=TRUE)



RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))


RhinoclemmysData$tablePDPercentage <- lapply(RhinoclemmysData$trees,
                                             FUN=PDindex,
                                             distribution=RhinoclemmysData$distribution,
                                             percentual=TRUE,
                                             root=TRUE)

RhinoclemmysData$matrixPDPercentage <- as.data.frame(
                                          matrix(
                                              unlist(RhinoclemmysData$tablePDPercentage),
                                              nrow= length(treeFiles),
                                              byrow=TRUE))


```


now, it is time to name trees, terminals and areas:

```{r}

RhinoclemmysData$nameTrees <- rownames(RhinoclemmysData$matrixPD) <-
                              rownames(RhinoclemmysData$matrixPDPercentage) <-
                                           gsub("Rhinoclemmys_","",
                                           gsub(".tre","",
                                                      treeFiles))

RhinoclemmysData$nameAreas <- colnames(RhinoclemmysData$matrixPD) <-
                              colnames(RhinoclemmysData$matrixPDPercentage) <-
                              rownames(RhinoclemmysData$distribution)

RhinoclemmysData$nameTerminals <- colnames(RhinoclemmysData$distribution)

```


now we can plot the trees

```{r}

for(treeNumber in 1:length(treeFiles)){

cat(RhinoclemmysData$nameTrees[[treeNumber]],"\n")

       plot.phylo(RhinoclemmysData$trees[[treeNumber]],
                  main=RhinoclemmysData$nameTrees[[treeNumber]])


   }

```

We perform some basic calculations:

```{r}


### killroy is here
### to do evaltree !!!!!!

# eval tree function

RhinoclemmysData$evalPD <- lapply(RhinoclemmysData$trees,
                                  FUN=evalTree,
                                  distribution=RhinoclemmysData$dist,
                                  root=TRUE)


# sum of deltas
RhinoclemmysData$sumDeltas <-   sapply(RhinoclemmysData$evalPD,
                                    function (x) {sum(x$abDelta/x$deltaPD)})


# max PD value

RhinoclemmysData$maxPD <- colnames(RhinoclemmysData$matrixPD)[apply(
                                      RhinoclemmysData$matrixPD,1,which.max)]




RhinoclemmysData$matrixPDPercentage

RhinoclemmysData$maxPD


```


Dependending on the tree/clock, PD prefers whether area B or F, but we do not know if the values in the analyses are close enough to consider the difference in PD values an artifact or a real difference given the tree / clock used.





```{r}



# swap branch lengths, allswap


RhinoclemmysData$swapBLalls <- lapply(RhinoclemmysData$trees,
                                  FUN=swapBL,
                                  distribution=RhinoclemmysData$dist,
                                  root=TRUE)


RhinoclemmysData$swapBLalls

RhinoclemmysData$swapBLunif <- lapply(RhinoclemmysData$trees,
                                  FUN=swapBL,
                                  distribution=RhinoclemmysData$dist,
                                  model = "uniform",
                                  root=TRUE)


RhinoclemmysData$swapBLunif


```

In this case, the tree selected changes the area selected, and we are confident in the results as in the three cases the branch swap does not affect the results.


And plotted into the trees:


```{r}

options(warn=-1)

for(treeNumber in 1:length(treeFiles)){

cat(RhinoclemmysData$nameTrees[[treeNumber]],"\n")


theTitle <- paste("Initial area selected:",RhinoclemmysData$evalPD[[treeNumber]]$InitialArea[1])



p0 <-    ggtree(RhinoclemmysData$trees[[treeNumber]], layout="rectangular", ladderize=TRUE,
                ##color=c("red","blue","red","blue","black","black","black"),
                color=c("black"),
                 size=0.5 ) +
         theme(legend.position="right") +
         labs(title = theTitle)



p <- p0 %<+% RhinoclemmysData$evalPD[[treeNumber]] +
             geom_tiplab(aes(color=areaDelta), size =3) +
             scale_colour_brewer("Area change", palette="Dark2")


print(p)


}
```

As this is too messy, we will focus our attention in areas B and F

```{r}

options(warn=-1)

for(treeNumber in 1:length(treeFiles)){

if(treeNumber == 1){tmpCol <- cbind(RhinoclemmysData$evalPD[[1]][,1],
                                 RhinoclemmysData$evalPD[[treeNumber]]$areaDelta)
                   }else{
                      tmpCol <- cbind(tmpCol,RhinoclemmysData$evalPD[[treeNumber]]$areaDelta)
                    }

}


rowNames <- tmpCol[,1]

tmpCol <- tmpCol[,-1]

colnames(tmpCol) <- RhinoclemmysData$nameTrees

row.names(tmpCol) <- rowNames

tmpCol

```


In all three cases _R._ _diademata_ changes the initial selection to the second "optional" area, what might suggests that for this species, its distribution (only in area B) or its branch length, could be leading the results.

To test our hypothesis, we can delete the species or we could remove the species or change its branch length to zero, or change its distribution to null, and rerun the analysis.

```{r}

## already done using evalTree; here only _R_diademata_ will be tested.

RhinoclemmysData$evalRdiademata <- lapply(RhinoclemmysData$trees,
                                   FUN=evalTerminal,
                                   distribution=RhinoclemmysData$distribution,
                                   tipToEval = "R_diademata",
                                   approach="lower",
                                   root=TRUE)

RhinoclemmysData$evalRdiademata


## removing the distribution of the species

RhinoclemmysData$distribution[,13] <- rep(0,14)

RhinoclemmysData$newPD <- lapply(RhinoclemmysData$trees,
                                  FUN=PDindex,
                                  distribution=RhinoclemmysData$dist,
                                  root=TRUE)


RhinoclemmysData$newMatrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$newPD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))

colnames(RhinoclemmysData$newMatrixPD) <- rownames(RhinoclemmysData$dist)

RhinoclemmysData$newMaxPD <- colnames(RhinoclemmysData$newMatrixPD)[apply(
                                      RhinoclemmysData$newMatrixPD,1,which.max)]


RhinoclemmysData$newMaxPD




```

As expected, now all trees give the same result, they select the area F. While this effect could be assigned to the distribution alone -the species is a singleton-. most of the species (17 out of 22) are singletons, therefore, the most plausible explanation must include the branch length.

```{r}

for(treeNumber in 1:length(treeFiles)){

##print(RhinoclemmysData$evalPD[[treeNumber]])

print(RhinoclemmysData$nameTrees[[treeNumber]])



finalResults <- RhinoclemmysData$evalPD[[treeNumber]]


### ordering according to terminalBL


finalResults <- finalResults[(order(finalResults$initialLength)),]

finalResults$shortLabelTerminal  <- sprintf("t%02d",
                                            1:length(RhinoclemmysData$trees[[treeNumber]]$tip.label))


plotResults <- ggplot(data=finalResults, aes(x= shortLabelTerminal, y= initialLength,
##plotResults <- ggplot(data=finalResults, aes(x= row.names(finalResults), y= initialLength,
                      shape="Actual",
                      colour=InitialArea)) +
               geom_point(size= 7) +
               geom_point(aes(x= shortLabelTerminal, y= lowerBranchLength,
##               geom_point(aes(x= row.names(finalResults), y= lowerBranchLength,
                              colour=lowerFinalArea,
                              shape="Lower_limit"), size=7) +
               labs(title = "Branch length change, lower limits.",
                    colour = "Area selected",
                    shape = "Terminal branch length value",
                    y = "Terminal branch length",
                    x = "Terminal")

print(plotResults)

plotResults <- ggplot(data=finalResults, aes(x= shortLabelTerminal, y= initialLength,
                      shape="Actual",
                      colour=InitialArea)) +
               geom_point(size= 7) +
               geom_point(aes(x= shortLabelTerminal, y= upperBranchLength,
                              colour=upperFinalArea,
                              shape="Upper_limit"), size=7) +
               labs(title = "Branch length change, upper limits.",
                    colour = "Area selected",
                    shape = "Terminal branch length value",
                    y = "Terminal branch length",
                    x = "Terminal")

print(plotResults)


}

```

To test whether internal branches have more impact than terminal branches, we can use the function evalTerminalvsInternal.


```{r}

data(tree)
 data(distribution)
 ## calculate values

 data1 <- evalTerminalvsInternal(tree,distribution,nTimes=10)
 plot1 <- lapply(data1,graficar)

## plot the results

 library(gridExtra)

 grid.arrange(
 plot1[[1]]+ggtitle(names(data1)[1]),
 plot1[[2]]+ggtitle(names(data1)[2]),
 plot1[[3]]+ggtitle(names(data1)[3]),
 plot1[[4]]+ggtitle(names(data1)[4]),
 nrow = 2)

```
