
---
title: "An empirical example: _Rhynoclemmys_ data."
author: "Daniel R. Miranda-Esquivel"
date: "2020 - 10 - 13"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---


# An empirical example: _Rhynoclemmys_ data.

We read the data, distribution and trees.


```{r}

options(warn=-1)
suppressMessages(library(ggtree))
#~ library(ape)
suppressMessages(library(blepd))
library(ggplot2)
#~ library(ggtree)



## read distributional data

setwd("../data/")

csvFile <- dir(pattern=".csv")


distribution <-          as.matrix(read.table(csvFile,
                                  stringsAsFactors=FALSE,
                                  header=TRUE,
                                  row.names=1,
                                  sep=",")
                          )

##qplot(distribution)



## read trees

treeFiles <- dir(pattern=".tre")

treeFiles

RhinoclemmysData <- list()

RhinoclemmysData$distribution <- distribution

RhinoclemmysData$trees  <-    lapply(treeFiles,FUN=read.tree)


distXY <- matrix2XY(distribution)



## plotting

## the trees


par(mfrow=c(2,1))

for( numero in 1:length(treeFiles)){

plotTree <-  ggtree(RhinoclemmysData$trees[[numero]], ladderize=TRUE,
                    color="black", size=2, linetype="solid") +
             geom_tiplab(size=3, color="black") +
             theme_tree2() +
             ggtitle(treeFiles[numero])


##
print(plotTree)

}



#~ Alternatively, we can plot the trees using ape

#~ for(treeNumber in 1:length(treeFiles)){
#~ 
#~        plot.phylo(RhinoclemmysData$trees[[treeNumber]],
#~                   main=treeFiles[numero])
#~ 
#~    }





## the distribution


par(mfrow=c(1,1))

plotDistrib <-ggplot(distXY,
               aes(x= Area, y=Terminal), size =11) +
              geom_point() +
               labs(title = "Distributions",
                    y = "",
                    x = "Area")

##
print(plotDistrib)

##cowplot::plot_grid(plotTree, plotDistrib, ncol=2)

# Using faceplot


#~ p2 <- facet_plot(plotTree, panel="dot", 
#~                  data=distXY, 
#~                  geom=geom_point(), 
#~                  mapping=aes(x= distXY$Area,y=distXY$Terminal), 
#~                  size =11, color='red3')


#~ print(p2)

##dev.off()

```


These two trees correspond to different data sources Molecular vs TotalEvidence data, and we want to test whether the data used will have any effect in the areas chosen, or if the area is selected given the terminal(s) branch length.



The branch length histogram shows there are two longer branches,  _R._ _aerolata_ and _R._ _rubida_ and these two are in area A, while area B is the richest.



```{r}

par(mfrow=c(2,1))

qplot(RhinoclemmysData$trees[[1]]$edge.length,bins=10,title=treeFiles[1])


qplot(RhinoclemmysData$trees[[2]]$edge.length,bins=10,title=treeFiles[2])

###dev.off()

```




Fist we calculate the PD value for each tree, we save the values as a table and later we convert the table to a matrix:

```{r}


RhinoclemmysData$tablePD <- lapply(RhinoclemmysData$trees,
                                   FUN=PDindex,
                                   distribution=RhinoclemmysData$distribution,
                                   root=TRUE)



RhinoclemmysData$matrixPD <- as.data.frame(
                                  matrix(
                                      unlist(RhinoclemmysData$tablePD),
                                      nrow= length(treeFiles),
                                      byrow=TRUE))


RhinoclemmysData$tablePDPercentage <- lapply(RhinoclemmysData$trees,
                                             FUN=PDindex,
                                             distribution=RhinoclemmysData$distribution,
                                             percentual=TRUE,
                                             root=TRUE)

RhinoclemmysData$matrixPDPercentage <- as.data.frame(
                                          matrix(
                                              unlist(RhinoclemmysData$tablePDPercentage),
                                              nrow= length(treeFiles),
                                              byrow=TRUE))



```

Now we test the efect of branch length swapping, first for terminal branch lengths.


```{r}


for( numero in 1:length(treeFiles)){
	
	for(modelo in c("simpleswap","allswap","uniform")){
		
		val <- swapBL(RhinoclemmysData$trees[[numero]],distribution,
		       model=modelo
		       )
		 
		 cat("\nTree=",treeFiles[numero],"\tModel=",modelo,"\n")      
		print.blepd(val)
		 cat("\n\n\n")
	}

}


```

For both trees, the terminal branch length is critical in the decision taken, if we swap all terminal branch lengths (= _allswap_) or if we replce them with a uniform distribution (_uniform_), the area selected change from A to B.

As the terminal branch lengths are distributed unequally, we might suspect that the resultds depeds heavily on the longest branches that inhabit the area A.


```{r}


## Trsting if we reduce the brach lenght form the actualsise to 0.0 using evalTerminal {lower}

testEvalTerminalALL <- evalTerminal(RhinoclemmysData$trees[[1]],distribution,tipToEval = "all",approach = "lower")

print.multiBlepd(testEvalTerminalALL)


## Trsting if we increase the brach lenght form the actualsise to the sum of all brach length  using evalTerminal {upper}


testEvalTerminalALL <- evalTerminal(RhinoclemmysData$trees[[1]],distribution,tipToEval = "all",approach = "upper")

print.multiBlepd(testEvalTerminalALL)


```




The area selected depends heavily in the branch length of _R._ _aerolata_ and _R._ _rubida_, if the brach length is just a little shorter, the area selected will change from area A to area B.


######



#~ now, it is time to name trees, terminals and areas:

#~ ```{r}

#~ RhinoclemmysData$nameTrees <- rownames(RhinoclemmysData$matrixPD) <-
#~                               rownames(RhinoclemmysData$matrixPDPercentage) <-
#~                                            gsub("Rhinoclemmys_","",
#~                                            gsub(".tre","",
#~                                                       treeFiles))

#~ RhinoclemmysData$nameAreas <- colnames(RhinoclemmysData$matrixPD) <-
#~                               colnames(RhinoclemmysData$matrixPDPercentage) <-
#~                               rownames(RhinoclemmysData$distribution)

#~ RhinoclemmysData$nameTerminals <- colnames(RhinoclemmysData$distribution)

#~ ```



#~ To test whether internal branches have more impact than terminal branches, we can use the function evalTerminalvsInternal.


#~ ```{r}

#~ data(tree)
#~  data(distribution)
#~  ## calculate values

#~  data1 <- evalTerminalvsInternal(tree,distribution,nTimes=10)
#~  plot1 <- lapply(data1,graficar)

#~ ## plot the results

#~  library(gridExtra)

#~  grid.arrange(
#~  plot1[[1]]+ggtitle(names(data1)[1]),
#~  plot1[[2]]+ggtitle(names(data1)[2]),
#~  plot1[[3]]+ggtitle(names(data1)[3]),
#~  plot1[[4]]+ggtitle(names(data1)[4]),
#~  nrow = 2)

#~ ```
