

## branch swap

An initial approach, is to evaluate the effect in PD when internal and terminal branch lengths are swapped. In this case it is not the sensitivity to the branch length as a parameter, but the stability to the actual branch lengths.

The function to perform the analysis is *swapBL*, that has four parameters: the tree, the distribution, the model to evaluate (valid models are "simpleswap", "allswap" -default value- and "uniform"), the number of times to swap (default value = 100), and branch to swap ("terminals" (default) or "internals").


Uning the default parameters we get.

```{r}

swapBL(tree = initialTree,
       distribution = dist4taxa)

```

As this is a tree where all branches are equal, there is no impact when the branch lengths are swapped.


Or we could use the random uniform branch length model.


```{r}

swapBL(tree = initialTree,
       distribution = dist4taxa,
       model = "uniform")

```

This is a tree where all branches are equal, therefore min and max are equal. There is no impact when the branch lengths are swapped, and areas A1A2 are selected.


