---
title: "Branch length evaluation for Phylogenetic Diversity: a worked example"
author: "Daniel R. Miranda-Esquivel"
date: "2020 - 10 - 11"
output:
  pdf_document:
    latex_engine: xelatex
---

# Four taxa and two areas

##  Preparing the data space

First, we load the required libraries:

```{r}

## cleaning
rm(list = ls())

## libraries

## installing and loading the package

##install.packages("../../blepd_0.mainRev.minorRev.tar.gz", repos = NULL, type="source")

library(blepd)

packageVersion("blepd")

## To plot trees you can use ggtree, ape or phytools. The example is based on
## ggtree as a matter of choice.

## More about ggtree
## https://guangchuangyu.github.io/ggtree-book/chapter-ggtree.html

## If you want to install ggtree go to
## https://bioconductor.org/packages/release/bioc/html/ggtree.html

library(ggplot2)

library(ggtree)

library(gridExtra)

library(RColorBrewer)

```

Now, we load the data included in the package: tree and distribution.

```{r}


#data(package = "blepd")

## trees

data(tree)

str(tree)

initialTree <- tree


## distributions

data(distribution)

str(distribution)

dist4taxa <- distribution

## distribution to XY to be able to plot the data, otherwise skip the next step

distXY <- matrix2XY(dist4taxa)


## plotting

## the tree

plotTree <-  ggtree(initialTree, ladderize=TRUE,
                    color="black", size=1, linetype="dotted") +
             geom_tiplab(size=6, color="black") +
             theme_tree2() +
             ggtitle("Four terminals, equal branch length")


##print(plotTree)



## the distribution

plotDistrib <- ggplot(data=distXY,
                      aes(x= Area, y= Terminal),
                      size =11) +
               geom_point() +
               labs(title = "Distributions",
                    y = "",
                    x = "Area")

##print(plotDistrib)

cowplot::plot_grid(plotTree, plotDistrib, ncol=2)

```

We check whether names in both objects: initialTree and dist4taxa, are the same.

```{r}

all(colnames(dist4taxa) == initialTree$tip.label)

```

We report the branch length, and calculate the PD values.

```{r}

initialTree$edge.length

initialPD <- PDindex(tree=initialTree, distribution = dist4taxa)

initialPD

```

As expected there is a tie between both areas.





## branch swap

An initial approach, is to evaluate the effect in PD when internal and terminal branch lengths are swapped. In this case it is not the sensitivity to the branch length as a parameter, but the stability to the actual branch lengths.

The function to perform the analysis is *swapBL*, that has four parameters: the tree, the distribution, the model to evaluate (valid models are "simpleswap", "allswap" -default value- and "uniform"), the number of times to swap (default value = 100), and branch to swap ("terminals" (default) or "internals").


Uning the default parameters we get.

```{r}

swapBL(tree = initialTree,
       distribution = dist4taxa)

```

As this is a tree where all branches are equal, there is no impact when the branch lengths are swapped.


Or we could use the random uniform branch length model.


```{r}

swapBL(tree = initialTree,
       distribution = dist4taxa,
       model = "uniform")

```

This is a tree where all branches are equal, therefore min and max are equal. There is no impact when the branch lengths are swapped, and areas A1A2 are selected.






##  Function to evaluate a single terminal

To test the effect of changing the branch length in a single terminal ("t1"), we will use the function **evalTerminal**. This function uses four parameters: tree, distribution, tipToEval (label of the tip), approach (two options: "lower"/"upper", to evaluate from 0 to the actual length or from the actual length to the sum of all branch lengths).

The function reports a S3 object, printable with **print.blepd**.

```{r}

terminalT1 <- evalTerminal(tree = initialTree,
             distribution = dist4taxa,
             tipToEval = "t1",
             approach = "lower" )

print.blepd(terminalT1)

terminalT1$delta

```

The lower limit reported when we change the branch length for terminal t1 is 0.99 [stored as *object**$delta**], therefore, this branch length will modify the area selected from A1A2 to A2, as the tie between the path between terminals t1/t3 (area A1) vs t2/t4 (area A2) will be solved in favour of t2/t4 when A1 is shorter.

In the same way, if we chage t1 to 1.01 the will break in favour of the area A1.


## Tree evaluation function

### branch length

The function to test all terminals at the same time is *evalTree*, with two parameters: the tree and the distribution. The function returns a data.frame object with 14 fields: labelTerminal, lowerBranchLength, InitialArea, lowerFinalArea, initialLength, upperBranchLength, upperFinalArea, changeLower, changeUpper, deltaUpper, deltaLower, deltaPD, areaDelta, and abDelta.


```{r}

finalResults <-   evalTree(tree = initialTree, distribution = dist4taxa)

## todo: errors to correct evalTree

finalResults

```


The extreme sensitivity of the PD results to the terminal branch length is seen in the column absolute length difference (=abDelta), as any length change -larger than 0-, will modify the area selected.

We plot the results to see the effect in each terminal, as a table.

```{r}

plotResults <- ggplot(data=finalResults, aes(x= labelTerminal, y= initialLength,
                      shape="Actual",
                      colour=InitialArea)) +
               geom_point(size= 7) +
               geom_point(aes(x= labelTerminal, y= lowerBranchLength,
                              colour=lowerFinalArea,
                              shape="Lower_limit"), size=7) +
               labs(title = "Branch length change, lower limits. Equal branches.",
                    colour = "Area selected",
                    shape = "Terminal branch length value",
                    y = "Terminal branch length",
                    x = "Terminal")

print(plotResults)

```

or plotted as a simple table.

```{r}

countFreqChanges <- table(finalResults$areaDelta)


countFreqChanges <- as.data.frame(countFreqChanges, ncol=1)


colnames(countFreqChanges) <- c("Area change","Freq")


row.names(countFreqChanges) <- NULL


countFreqChanges


```

or plotted into the tree:

```{r}


theTitle <- paste("Initial area selected:",finalResults$InitialArea[1])

p0 <-    ggtree(initialTree, layout="slanted", ladderize=TRUE,
                color=c("red","blue","red","blue","black","black","black"),
                 size=0.8 ) +
         theme(legend.position="right") +
         labs(title = theTitle)


p <- p0 %<+% finalResults + geom_tiplab(aes(color=areaDelta), size =6) +
          scale_colour_brewer("Area change", palette="Dark2")


print(p)


```


For terminals t1/t3, a change from 1 to 0.99 in branch length -the lower limit (=L)- will change the initial area selected (A1A2) to A2; or a change from 1 to 1.01 in branch length -the upper limit(=U)-, will change the area to A1.

